---
title: "Example Exam 3"
author:   
  - Jinzhe Yang
date: last-modified
format: 
  html: default
  pdf:
    keep-tex: true
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
         \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
toc: true
language: 
  title-block-author-single: " "
toc-title-document: "Content"
crossref-fig-title: "Figure"
theme: lumen
title-block-banner-color: Primary
callout-warning-caption: "Warning"
callout-note-caption: "Note!"
callout-tip-caption: "Tip"
editor: visual
---

```{r}
## helper-codes

# install.packages("mvtnorm") # uncomment if this package is not installed.
library(mvtnorm)

######## Problem 1
load("cauchy.RData") # loads the data vector x.

# Defines the log-Cauchy density
dcauchy <- function(x, mu) {
    return(-log(pi) - log(1 + (x - mu)^2))
}

########  Problem 2
# Nothing to load for this problem

########  Problem 3
load("./cauchy.RData") # loads the data vector x.

######## Problem 4

load("./debtratio.RData") # load the vector y and matrix of covariates X

```

## 1. Cauchy distribution
### a. Posterior Distribution
```{r}
LogPost <- function(x, theta, mu_0 = 0, sigma2_0 = 1) {
    logPrior <- dnorm(theta, mean = mu_0, sd = sqrt(sigma2_0), log = TRUE)

    logLikeli <- sum(dcauchy(x, mu = theta))

    return(logPrior + logLikeli)
}
thetaGrid <- seq(-10, 10, length = 1000)
logPostList <- rep(NA, length(thetaGrid))
for (i in 1:length(thetaGrid)) {
    logPostList[i] <- LogPost(x, thetaGrid[i])
}
# logPostList
LogPost(x, -9.97997998)
postList <- exp(logPostList)
postDen <- postList / sum(postList) / (thetaGrid[2] - thetaGrid[1])
plot(thetaGrid, postDen,
    type = "l",
    ylab = "posterior density"
)
```

### b. Posterior Mean
Under quadratic loss function, the optimal point estimate is posterior mean. Use intergration to calculate posterior mean

```{r}
sum(thetaGrid * postDen * (thetaGrid[2] - thetaGrid[1]))
```

```{r}
# check posterior mode
initialValue <- 0
optim(initialValue, LogPost,
    method = "BFGS", control = list(fnscale = -1),
    hessian = TRUE,
    x = x
)

# check total posterior pdf intergrated to 1
sum(postDen * (thetaGrid[2] - thetaGrid[1]))
```

### c. Construct at 95% credible interval for θ.
HPD or equal-tail. Using equal-tail, 2.5% area on the left tail, 2.5% area on the right tail.

```{r}
cumulative_list <- rep(NA, length(thetaGrid))
for (i in 1:length(thetaGrid)) {
    area <- postDen[i] * (thetaGrid[2] - thetaGrid[1])
    if (i == 1) {
        cumulative_list[i] <- area
    } else {
        cumulative_list[i] <- area + cumulative_list[i - 1]
    }
}
# cumulative_list
# which(cumulative_list == 0.025 )
binSize <- thetaGrid[2] - thetaGrid[1]
cumulative_list[1]

# cumsum function do cdf

tail(thetaGrid[cumsum(postDen * binSize) <= 0.025], 1)
tail(thetaGrid[cumsum(postDen * binSize) <= 0.975], 1)

head(thetaGrid[cumsum(postDen * binSize) >= 0.975], 1)
```

## 2. Binomial data

## 3. Metropolis-Hastings
```{r}
# random walk metropolis
# return samples and accpetance rate
RandomWalk <- function(LogPost, mu_0, c, nSim, nBurn, ...) {
    draws <- rep(NA, nSim)
    acceptance <- NA
    rejNum <- 0
    for (i in 1:nSim) {
        if (i == 1) {
            # first draw
            draws[i] <- rnorm(1, mean = mu_0, sd = sqrt(c))
        } else {
            temp <- rnorm(1, mean = draws[i - 1], sd = sqrt(c))
            u <- runif(1)
            alpha <- min(1, exp(LogPost(theta = temp, ...) - LogPost(theta = draws[i - 1], ...)))
            if (u > alpha) {
                # reject
                draws[i] <- draws[i - 1]
                rejNum <- rejNum + 1
            } else {
                draws[i] <- temp
            }
        }
    }
    return(list(draws = draws[(nBurn + 1):nSim], acceptance = rejNum / nSim))
}

res <- RandomWalk(LogPost, mu_0 = 0, c = 0.04, nSim = 1100, nBurn = 100, x)
head(res$draws)
res$acceptance
```

```{r}
# use the code from hw2
source("RWMsampler.R")

initVal = c(0)
Sigma = matrix(1,1,1)
c = 0.04
rmvnorm(1, mean = initVal, sigma = c * Sigma)
draws = RWMsampler(LogPost, c(0), nSim = 1100, nBurn = 100, Sigma = matrix(1,1,1), c = 0.04, x)
mean(draws)
```

### b. 
show sample mean
```{r}
draws <- res$draws
sampleMeanList <- rep(NA, length(draws))
for (i in 1:length(draws)) {
    sampleMeanList[i] <- mean(draws[1:i])
}

plot(1:length(sampleMeanList), sampleMeanList, type = "l", xlab = "sample size")

sampleMeanList[length(draws)]
mean(draws)
```

## 4. Beta regression
```{r}
LogPostBetaReg <- function(Beta, X, y, mu_0 = 0, sigma2_0 = 10^2) {
    mu <- exp(X %*% Beta) / (1 + exp(X %*% Beta))
    fi <- 4
    logLikeli <- sum(dbeta(y, mu * fi, (1 - mu) * fi, log = TRUE))
    logPrior <- sum(dnorm(Beta, mean = mu_0, sd = sqrt(sigma2_0), log = TRUE))

    # logPrior <- dmvnorm(Beta, mean = rep(mu_0, 5), sigma = sigma2_0*diag(5), log = TRUE)

    return(logLikeli + logPrior)
}

initialValue <- c(0, 0, 0, 0, 0)
LogPostBetaReg(initialValue, X, y) # (1,2,3,4,5) returns -inf

res <- optim(initialValue, LogPostBetaReg,
    method = "BFGS", control = list(fnscale = -1), hessian = TRUE,
    X = X,
    y = y
)

postMode <- res$par
postCov <- -solve(res$hessian)
postMode
postCov
```

### b.
Check the marginal distribution for `logsale`
```{r}
thetaGrid <- seq(-0.1, 0.2, length = 1000)
densityList <- dnorm(thetaGrid, mean = postMode[4], sd = sqrt(postCov[4, 4]))
plot(thetaGrid, densityList, type = "l")

lower_95 <- qnorm(0.025, mean = postMode[4], sd = sqrt(postCov[4, 4]))
upper_95 <- qnorm(0.025, mean = postMode[4], sd = sqrt(postCov[4, 4]), lower.tail = FALSE)

abline(v = lower_95, col = "green", lty = 3, lw = 2)
abline(v = upper_95, col = "green", lty = 3, lw = 2)
```

$\beta_4 = 0$ is covered in 95% credible interval. So covariate `logsale` can be removed.

### c.
```{r}
# get simulation draws from normal approximation
draws <- rmvnorm(1000, mean = postMode, sigma = postCov)
x_tilde <- c(1, 0.5, 1, 5, 0)
mu <- exp(draws %*% x_tilde) / (1 + exp(draws %*% x_tilde))
fi <- 4
y_tilde <- rbeta(length(mu * fi), mu * fi, (1 - mu) * fi)
hist(y_tilde, col = "lightgray", freq = FALSE)
```